{\scshape Denicek}: Computational Substrate for\\ Document-Oriented End-User Programming

### 2025-03-30 paper.tex
Case study implementation

\begin{figure*}[t]
\includegraphics[width=0.98\textwidth,clip,trim=0cm 1cm 1cm 0.5cm]{fig/datnicek.pdf}
\caption{A notebook visualizing Air traffic accident data from Eurostat. The user loads data in
a code cell (1), edits it in a grid that infers edit operations via programming by demonstration
(2), turned to a code cell (3) and then visualized.}
\label{fig:datnicek}
\end{figure*}

@@

Denicek is intended as a low-level computational substrate that can be used as the basis for
interactive programming systems that view programs as documents. The most prominent example
of such systems today are notebook environments for data science. We developed Datnicek, a
simple notebook system that shows the ability of Denicek to support rich interactive user
experiences. In this section, we present Datnicek and reflect on its development. A simple data
exploration conducted in Datnicek is shown in Fig.~\ref{fig:datnicek}.

@@

\begin{itemize}
\item \emph{Structure Editing.} Code in code cells should be edited through structure editor
  as in Histogram \cite{petricek-2019-histogram}. This allows Datnicek to keep track of the code
  structure and potentially makes the system accessible to non-programmers \cite{mcnutt-2023-projectional}.
  We do not aim to implement advanced keyboard-based editing \cite{moon-2022-tylr,beckman-2023-sandblocks}.

@@

\item \emph{Wrangling Code Synthesis.} Interactively constructed data transformations should be
  turned into code that can be checked and further edited as in Wrex \cite{drossos-2020-wrex}.
\end{itemize}

@@

\subsection{Implementation}

data representation

(maybe show fig with source)

operations

@@

>>>>

\begin{figure*}[t]
\vspace{-0.5em}
\includegraphics[width=0.98\textwidth,clip,trim=0cm 1cm 1cm 0.5cm]{fig/datnicek.pdf}
\vspace{-0.5em}
\caption{A notebook visualizing Air traffic accident data from Eurostat. The user loads data in
a code cell (1), edits it in a grid that infers edit operations via programming by demonstration
(2), turned to a code cell (3) and then visualized.}
\label{fig:datnicek}
\vspace{-0.5em}
\end{figure*}

@@

Denicek is a low-level computational substrate. It serves best as the basis for interactive
programming systems that view programs as documents. The most prominent example of such systems
today are notebook environments for data science. To explore this use case, we have developed
Datnicek, a simple notebook system that shows the ability of Denicek to support rich interactive user
experiences. In this section, we present Datnicek and reflect on its development. A simple data
exploration conducted in Datnicek is shown in Fig.~\ref{fig:datnicek}.

@@

\begin{itemize}
\item \emph{Structure Editing.} Code in code cells should be edited through structure editor
  as in Histogram \cite{petricek-2019-histogram}. This allows Datnicek to keep track of the code
  structure and potentially makes the system accessible to non-programmers \cite{mcnutt-2023-projectional}.
  We do not yet aim to implement advanced keyboard-based editing \cite{moon-2022-tylr,beckman-2023-sandblocks}.

@@

\item \emph{Wrangling Code Synthesis.} Interactively constructed data transformations should be
  turned into code that can be checked by the data scientist and further edited as in Wrex \cite{drossos-2020-wrex}.
\end{itemize}

@@

\subsection{Implementation}
Datnicek is implemented using the Elm architecture \cite{fowler-2020-mvu},
where a system maintains a state that is updated through events. The state consists
of the list of Denicek edits, transient state of the user interface, current computed document
and other cached information. All interactions that affect the notebook
use an event that appends edits to the list of Denicek edits. Remaining events
update the transient user interface state. The support for collaborative editing has been
implemented through an ad-hoc mechanism.

\subsubsection*{User Interface}
Datnicek notebooks consists of a series of cells. In the code cells, the user can
use the ``+'' button to add a new variable, add an operation to the end of an existing chain
of operations, but also to specify an argument of an operation. The button opens
a menu showing all valid options based on the current context.

In the interactive grid, the user can edit column headers and table cells. When they edit
a value, Datnicek suggests generalised edits to be applied to the entire
column or table. For example, when the user changes the ``0 p'' value to ``0'', Datnicek
suggests to replace `` p'' with the empty string in all cells. Other suggested edits include renaming
or deleting a column, splitting a column using a delimiter and filtering rows based on the selected
value. Datnicek can also turn edits performed interactively into a new code cell.

\subsubsection*{Programmatic Code Cells}
A Denicek document representing a Datnicek notebook is shown in Fig.~\ref{fig:source}. A
notebook is a record storing individual cells as fields. The tag determines the type of the cell.

The programming language used in code cells is inspired by the data exploration calculus \cite{petricek-2020-live}.
A cell consists of a sequence of bindings of the form $\textnormal{\ident{let}}~v=e$ that assign
an expression $e$ to a variable $v$. An expression can be a reference to a variable,
global value (such as \ident{data} and \ident{vis}) or a method invocation $e.m(e_1, \ldots, e_n)$.

Bindings are represented as fields of a record and expressions use the formula representation
discussed in \S\ref{sec:impl-eval}. Operations and global values are identified by an absolute
reference pointing to the special \ident{/\$datnicek} namespace. Chains of method calls are
represented using nested formulas. Evaluation of formulas proceeds as in Webnicek
and wraps {\small\Verb_<x-formula>_} in {\small\Verb_<x-evaluated>_}.
The user interface displays the original formula alongside with the evaluated result, supporting
tables and Compost data visualizations \cite{petricek-2021-compost}. Edits to the formula
invalidate dependent results as discussed in \S\ref{sec:impl-incremental}.

\subsubsection*{Interactive Grid Cells}
Grid cells consist of a reference to the data source they edit and a collection of
{\small\Verb_<transform>_} nodes that represent individual data transformations constructed via
programming by demonstration. A transformation consists of metadata,
a sequence of edits and a formula representing the transformation (with {\small\Verb_<x-hole>_}
standing for the data table to be transformed).

A transformation may correspond to multiple underlying Denicek edits. For example, splitting
``foo/bar'' using ``/'' as the delimiter adds a new column ``bar'', copies values from
``foo/bar'', renames ``foo/bar'' to ``foo'' and then transforms the values in the new columns
by dropping the part before and after the delimiter, respectively.

Interactive grid also supports conditional transformations that affect rows satisfying a given
condition (e.g., deleting rows with a specific value in a given column). As discussed
in \S\ref{sec:system-defs}, Denicek does not allow specifying conditional selectors.
To support conditional transformations, we add a special kind of (virtual) edit
{\small\Verb_<x-expandable-edit>_} that stores the underlying edit and a condition.
As in \S\ref{sec:impl-interaction}, when applying the edit, Datnicek finds all rows for which the
condition holds and generates a single non-conditional edit for each such row.

@@

edits and formula in interactive should be unified
